---
title: "Accessibility Map Example - Burkina Faso"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "s5joschi"
date: "April 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installation
In order to use the package you need to have GRASS GIS installed. For installation please refer to: https://grass.osgeo.org/download/. 

The easiest way to install the Accessibility package is using the *devtools* package.  The installation command should look like this: **devtools::install_github("s5joschi/accessibility")**.  Activate the library as usual before usage **library(AccessibilityMaps)**. 


# Setup Working Directory and get data
The following code creates a working directory, input and output folders and downloads the example data. 
```{r setup wd and data}


```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
