---
title: "Accessibility Map Example - Burkina Faso"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "s5joschi"
date: "April 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installation
In order to use the package you need to have GRASS GIS installed. For installation please refer to: https://grass.osgeo.org/download/. 

The easiest way to install the Accessibility package is using the *devtools* package.  The installation command should look like this: 
```{r installation}
devtools::install_github("s5joschi/accessibility")
library(AccessibilityMaps)
```

# Setup Working Directory and get data
The following code creates a working directory, input and output folders and downloads the example data. 
```{r setup wd and data}
dir.create("accessibility_example")
setwd("accessibility_example/")

download.file(
  "https://github.com/s5joschi/accessibility/raw/master/example/input.zip",
  destfile = "input.zip"
)
unzip("input.zip")

dir.create("output")

```

# load and reprocess data
We load data for Burkina Faso. 
```{r load}
# landcover 
r_landcover<-raster("input/landuse/esa-copernicus-prototype_0005dergrees.tif")
# digital elevation model (DEM)
r_dem<-
  raster(
    "input/dem/bf_elevation.tif"
    )
# administrational boundaries
spodf_admin<-
  readOGR("input/admin/",
          "gis_osm_places_a_free_1"
          )
# OSM roads
spodf_roads<-
  readOGR(
    "input/roads/",
    "gis_osm_roads_free_1_mainroads"
    )
# OSM places
spodf_sources<-
  readOGR(
    "input/admin/",
    "gis_osm_places_free_1"
    )

# plot some of the base-data
plot(r_landcover)
plot(spodf_admin,border="red",add=T)
plot(spodf_roads,col="black",add=T)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
